---
title: "Employment Analysis"
author: "Leala Darby"
date: "03/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
First load all required packages:
```{r message=FALSE, warning=FALSE}
library(car)
library(tseries)
```

Load in the data:
```{r}
dat <- read.csv("employment_data.csv", fileEncoding = 'UTF-8-BOM')
head(dat)
```

Create a time series object from the data and plot
```{r}
ts_dat_test <- ts(dat[, 2], start = c(1978, 2), end = c(2020, 8), frequency = 12)
plot.ts(ts_dat_test)
abline(v = 1993, col = "blue")
abline(v = 2020, col = "blue")
```
Instructed to truncate data from January 1993 to December 2019 (inclusive)
```{r}
dat[dat$Observation.times == "Jan-93",]
dat[dat$Observation.times == "Dec-19",]
```
So we only need rows 180-503.
```{r}
trunc_dat <- dat[180:503,]
ts_dat <- ts(trunc_dat[, 2], start = c(1993, 1), end = c(2019, 12), frequency = 12)
plot.ts(ts_dat)
plot.ts(diff(ts_dat))
```
The trend in mean is readily observable.
Difficult to determine a trend in variance - there appears to be frequent changes, which are easier to see after incorporating lags of 1.

As a rough test (Levene's isn't really valid because time series data isn't independent)
```{r}
length(ts_dat)
Group <- c(rep(1,81), rep(2, 81), rep(3, 81), rep(4, 81))
leveneTest(ts_dat, Group)
```
The small p-value of $0.0001013$ confirms that the data exhibits heteroscedasticity. Therefore we will perform a log transformation to attempt to reduce this:
```{r}
log_ts_dat <- log(ts_dat)
plot.ts(cbind(ts_dat, log_ts_dat))
leveneTest(log_ts_dat, Group)
```

At a significance level of 5%, the p-value above of 0.2245 provides very weak evidence and we fail to reject the null hypothesis of equal variance among groups. Thus the heteroscedasticity has been reduced.

Next, to reduce the trend in mean, apply diff to incorporate lags of 1 to our TS with stabilised variance:
```{r}
f_ts_dat <- diff(log_ts_dat, 1)
plot.ts(cbind(ts_dat, log_ts_dat, f_ts_dat))
```
To confirm constant mean and variance and a Gaussian distribution for $u_t$, a Shapiro-Wilk normality test is performed:
```{r}
hist(f_ts_dat)
shapiro.test(f_ts_dat)
```
The small p-value indicates likely non-normality, but this test isn't really valid for TS. Instead, check statistically for stationarity using the Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test:
```{r}
kpss.test(ts_dat)
kpss.test(log_ts_dat)
kpss.test(f_ts_dat)
```
The final ts has a high p-value of 0.1, which is statistically significant at a significance level of 5%. Therefore we fail to reject the null hypothesis, and have reasonable evidence that the final ts is stationary.




















